---
title: "Example Species Report"
author: "Peter Solymos"
output: rmarkdown::html_vignette
params:
  species: "AlderFlycatcher"
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
library(cure4insect)
library(knitr)
library(rgdal)
set_options(verbose=0)
#set_options(path = "d:/abmi/reports")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

load_common_data()

species <- params$species
info <- as.list(get_species_table()[species, c("SpeciesID", "CommonName", "ScientificName", "TSNID")])
info <- lapply(info, as.character)
info[["SpeciesID"]] <- paste0("`\"", info[["SpeciesID"]], "\"`")
info[["ScientificName"]] <- paste0("_", info[["ScientificName"]], "_")
names(info) <- c("Species ID", "Common name", "Scientific name", "TSN ID")
```

# Introduction

This example report has been produced using the
**cure4insect** [R](https://www.r-project.org/) package
which a decision support tool that provides an interface to enable
custom reporting for intactness and sector effects
based on estimates and predictions created by the [Alberta
Biodiversity Monitoring Institute (ABMI)](http://abmi.ca/)
in collaboration with the
[Boreal Avian Modelling (BAM) Project](http://www.borealbirds.ca/).

```{r settings,results='asis',echo=FALSE}
for (i in 1:length(info))
    cat("* ", names(info)[i], ": ", info[[i]], "\n", sep="")
```

# Relative abundance by land cover types

## Habitat associations in northern Alberta

Models were used to show how species' relative abundance differed among
vegetation and human footprint types in the Boreal and Foothills regions of
Alberta. Predicted species abundance in each habitat type is shown with bars.
Vertical lines indicate 90% confidence intervals.
Dots within the forested habitat types show predicted species abundance in
cutblocks of various ages.

```{r hab-veg,echo=FALSE,fig.height=6,fig.width=12,out.width='100%'}
hab_veg <- plot_abundance(species, type="veg_coef", plot=FALSE)
hab_soil_tr <- plot_abundance(species, type="soil_coef", paspen=1, plot=FALSE)
hab_soil_nt <- plot_abundance(species, type="soil_coef", paspen=0, plot=FALSE)
ylim <- c(0, max(hab_veg, hab_soil_nt, hab_soil_tr))
plot_abundance(species, type="veg_coef", ylim=ylim)
```

## Habitat associations in southern Alberta

Models were used to show how species, relative abundance differed among soil
and human footprint types in the grassland and parkland regions of
Alberta. Due to natural disturbances and vegetation succession,
varying amounts of aspen and other trees may be present on each soil type;
the presence/absence of trees greatly affects the presence and abundance of
many biota. As such, information is presented twice for southern Alberta:
once for sites with trees present, and once for sites without trees present.
Predicted species abundance in each soil/human footprint type is shown with bars.
Vertical lines indicate 90% confidence intervals.

```{r hab-soil,echo=FALSE,fig.height=6,fig.width=12,out.width='100%'}
op <- par(mfrow=c(1,2))
plot_abundance(species, type="soil_coef", paspen=1, ylim=ylim, main="Treed sites")
plot_abundance(species, type="soil_coef", paspen=0, ylim=ylim, main="Non-treed sites")
par(op)
```

## Linear footprint relationships

For linear footprint in northern Alberta.
The pairs of points show the change in relative abundance expected between the
average habitat with no linear footprint (left point) and average habitat when
10% linear footprint is present (right point).

```{r hab-lin,echo=FALSE,fig.height=6,fig.width=12,out.width='100%'}
op <- par(mfrow=c(1,2))
plot_abundance(species, type="veg_lin", main="Northern Alberta")
plot_abundance(species, type="soil_lin", main="Southern Alberta")
par(op)

```

# Predicted relative abundance throughout Alberta

```{r spp-data,results='hide',echo=FALSE,warning=FALSE,message=FALSE}
y <- load_species_data(species)
r <- rasterize_results(y)
subset_common_data(id=NULL, species=species)
x <- calculate_results(y)

fn <- paste0("http://sc-dev.abmi.ca/reports/",
    getOption("cure4insect")$version, "/misc/raw_all.rda")
con <- url(fn)
load(con)
close(con)

z <- do.call(rbind, lapply(res, flatten))
class(z) <- c("c4idf", class(z))
```

## Relative abundance and prediction uncertainty

Habitat association models, plus models describing how species varied
spatially and with climate gradients were used to predict species abundance in
1 km$^2$ spatial units under reference conditions.
Predictions of relative abundance of the species in each 1 km$^2$ unit were made
based on the vegetation and human footprint present in the 1 km$^2$ unit in
(_current condition_) and
after all human footprint in the 1 km$^2$ unit had been 'backfilled' based on native
vegetation in the surrounding area (_reference condition_).
Pixels depicted in red are predicted to
have the highest abundance for the species, grading through light tan to dark
blue where the species is predicted to be less abundant or absent.

To highlight the degree of uncertainty in the models,
we estimated the prediction _standard error_ and _coefficient of variation_
for each 10 km x 10 km units based on bootstrap predictions of current abundance.
10 km x 10 km units depicted in red have the highest standard error,
while 10 km x 10 km units in dark green have the lowest.

```{r raster-plots,echo=FALSE,fig.height=12,fig.width=8,out.width='100%'}

col1 <- colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#E0F3F8","#91BFDB","#4575B4")))(100)
col2 <- colorRampPalette(rev(c("#A50026", "#D73027", "#F46D43", "#FDAE61", "#FEE08B", "#D9EF8B",
    "#A6D96A", "#66BD63", "#1A9850", "#006837")))(100)
op <- par(mfrow=c(2,2), mar=c(1,1,1,1))
plot(sqrt(r[["NC"]]), col=col1, axes=FALSE, box=FALSE, main="Current abundance")
plot(sqrt(r[["NR"]]), col=col1, axes=FALSE, box=FALSE, main="Reference abundance")
plot(r[["SE"]], col=col2, axes=FALSE, box=FALSE, main="Stadard error (current)")
plot(r[["CV"]], col=col2, axes=FALSE, box=FALSE, main="Coef. of variation (current)")
par(op)
```

## Species intactness

Intactness compares the current and reference abundances in each pixel.

For each 1 km$^2$ unit the % ratio between predicted current and reference
conditions was determined. In 1 km$^2$ units depicted in green the species
was predicted to have higher abundance under present conditions than under
reference conditions, with the opposite true for 1 km$^2$ units depicted in pink.
The intensity of green and pink depict the relative magnitude of increase or
decrease for the species between reference and current conditions.

```{r raster-si2,echo=FALSE,fig.height=6,fig.width=4,out.width='100%'}
col3 <- colorRampPalette(c("#C51B7D","#E9A3C9","#FDE0EF","#E6F5D0","#A1D76A","#4D9221"))(100)
op <- par(mfrow=c(1,1), mar=c(1,1,1,1))
plot(r[["SI2"]], col=col3, axes=FALSE, box=FALSE, main="Species intactness")
par(op)
```

The table shows provincial abundance totals and corresponding intactness values:

```{r intactness-table,echo=FALSE,results='asis'}
df1 <- data.frame(x$intactness)
colnames(df1)[2:3] <- c("Lower", "Upper")
kable(df1, digits=2)
```


## Sector effects

The effect of each type of human development (herein called a sector)
on a species relative abundance is the product of the area of the sector's
footprint, and the average "per unit area" effect of that sector's footprint
on the species relative abundance (_unit effect_). The combination of the two
results in the regional total effect on expected abundance of a species (_regional total effect_).
When the effect on species is only quantified within the sector footprint,
i.e. not taking into account native vegetation, we call it _under the footprint_ sector effect.


```{r secor-plot,echo=FALSE,fig.height=12,fig.width=12,out.width='100%'}
op <- par(mfrow=c(2,2))
plot_sector(x, type="regional", main="Regional")
plot_sector(x, type="underhf", "Under HF")
plot_sector(x, type="unit", "Unit")
plot.new()
par(op)
```

The table shows the sector effect results for the species across the entire province:

```{r sector-table,echo=FALSE,results='asis'}
df2 <- data.frame(x$sector)[,-1]
kable(df2, digits=2)
```

We can also look at where does the species fall among the rest of the species
in terms of the sector effects, $\times$ marks the species:

```{r secor-multi,echo=FALSE,fig.height=12,fig.width=6,out.width='100%'}
sectors <- c("Agriculture","Forestry","Energy","RuralUrban","Transportation")
h <- function(tmp) {
    v <- tmp[species, sectors]
    pmax(-100, pmin(100, v))
}
op <- par(mfrow=c(3,1))
tmp <- plot_sector(z, type="regional", main="Regional")
points(1:5, h(tmp), pch=4, col=1, cex=3)

tmp <- plot_sector(z, type="underhf", main="Under HF")
points(1:5, h(tmp), pch=4, col=1, cex=3)

tmp <- plot_sector(z, type="unit", main="Unit", ylim=c(-200,200))
points(1:5, h(tmp), pch=4, col=1, cex=3)
par(op)
```

# Regional summaries

The region of interest can be defined based for example on a spatial polygon.
Based on the spatial subset, we summarize abundance, intactness, and sector effects.

```{r spid,echo=FALSE,results='hide'}
ply <- readOGR(dsn=system.file("extdata/OSA_bound.geojson", package="cure4insect"))
ID <- overlay_polygon(ply)
## write IDs into a text file
#write.table(data.frame(SpatialID=ID), row.names=FALSE, file="SpatialID.txt")
AB <- readOGR(dsn=system.file("extdata/AB_bound.geojson", package="cure4insect"))
AB <- spTransform(AB, proj4string(r))
ply <- spTransform(ply, proj4string(r))
## calculate regional stats
subset_common_data(id=ID, species)
xreg <- calculate_results(y)
## clip raster
rreg <- crop(r, ply)
rreg <- mask(rreg, ply)
```

```{r plot-region,echo=FALSE,fig.height=6,fig.width=4,out.width='100%'}
par(mar=c(1,1,1,1))
plot(AB, col="grey", border="grey")
plot(ply, col="tomato", border="tomato", add=TRUE)
par(op)
```

```{r raster-plots-reg,echo=FALSE,fig.height=6,fig.width=8,out.width='100%'}
op <- par(mfrow=c(1,2), mar=c(1,1,1,1))
plot(sqrt(rreg[["NC"]]), col=col1, axes=FALSE, box=FALSE, main="Current abundance in OSA")
plot(sqrt(rreg[["NR"]]), col=col1, axes=FALSE, box=FALSE, main="Reference abundance in OSA")
par(op)
```

```{r intactness-table-reg,echo=FALSE,results='asis'}
df3 <- data.frame(xreg$intactness)
colnames(df3)[2:3] <- c("Lower", "Upper")
kable(df3, digits=2)
```

```{r sector-table-reg,echo=FALSE,results='asis'}
df4 <- data.frame(xreg$sector)[,-1]
kable(df4, digits=2)
```

```{r secor-plot-reg,echo=FALSE,fig.height=12,fig.width=12,out.width='100%'}
op <- par(mfrow=c(2,2))
plot_sector(xreg, type="regional", main="Regional in OSA")
plot_sector(xreg, type="underhf", "Under HF in OSA")
plot_sector(xreg, type="unit", "Unit in OSA")
plot.new()
par(op)
```
