---
title: 'cure4insect tutorial'
author: "Peter Solymos"
date: "2018-06-05"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: false
documentclass: article
fontsize: 11pt
classoption: letterpaper
urlcolor: blue
---

```{r global_options,include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

## What is cure4insect?

**cure4insect** is short for **Cu**stom **Re**porting **for**
**In**tactness and **Sect**or Effects. It is an
[R](https://www.r-project.org/) package and is a decision support tool
that provides an interface to enable
custom reporting for intactness and sector effects
based on estimates and predictions created by the [Alberta
Biodiversity Monitoring Institute (ABMI)](http://abmi.ca/)
in collaboration with the
[Boreal Avian Modelling (BAM) Project](http://www.borealbirds.ca/).

See some examples produced by the package:

* [Few slides about the motivations](https://abbiodiversity.github.io/cure4insect/intro/)
* [Example species report](https://abbiodiversity.github.io/cure4insect/example/)
* [Custom report](https://abbiodiversity.github.io/cure4insect/site/)
* [Web app](http://sc-dev.abmi.ca/ocpu/apps/ABbiodiversity/cure4insect/www/)

## License

The estimates, predictions, and related documentation are &copy; ABMI (2014&ndash;2018) under a [CC BY-SA 4.0 license](http://creativecommons.org/licenses/by-sa/4.0/).

The R package itself is licensed under [MIT license](https://github.com/ABbiodiversity/cure4insect/blob/master/LICENSE.md) &copy; 2018 Peter Solymos & ABMI.

## Install

Only GitHub version available now. If you have trouble installing the package,
please file an [issue](https://github.com/ABbiodiversity/cure4insect/issues).

```{r}
devtools::install_github("ABbiodiversity/cure4insect")
```

Now we can load the package:

```{r}
library(cure4insect)
```

## Workflow with 1 species

`id` is a vector of `Row_Col` type IDs of 1 km<sup>2</sup> pixels,
`species` is a vector of species IDs:

```{r}
load_common_data()

## define spatial and species IDs (subsets)
Spp <- "Ovenbird"
ID <- c("182_362", "182_363", "182_364", "182_365", "182_366", "182_367",
    "182_368", "182_369", "182_370", "182_371", "182_372")

subset_common_data(id=ID, species=Spp)
## check subsets
str(get_subset_id())
str(get_subset_species())

## load species data
y <- load_species_data(Spp)

## calculate results and flatten to a 1-liner
x <- calculate_results(y)
x
flatten(x)
```

## Spatial subset specifications

All the possible spatial IDs can be inspected as:

```{r}
str(get_all_id())
plot(xy <- get_id_locations(), pch=".")
summary(xy)
```

Spatial IDs can be specified as planning/management regions:

```{r}
## Natural Regions
ID <- get_all_id(nr=c("Boreal", "Foothills"))
## Natural Subregions
ID <- get_all_id(nsr="Lower Boreal Highlands")
## Land Use Framework regions
ID <- get_all_id(luf="North Saskatchewan")
```

Alternatively, `id` can refer to quarter sections
using the `MER-RGE-TWP-SEC-QS` format:

```{r}
Spp <- "Ovenbird"
QSID <- c("4-12-1-2-SE", "4-12-1-2-SW", "4-12-1-3-SE", "4-12-1-3-SW")
qs2km(QSID) # corresponding Row_Col IDs
```

The `subset_common_data` function recognizes `MER-RGE-TWP-SEC-QS` type
spatial IDs and onvert those to the `Row_Col` format using
the nearest 1 km<sup>2</sup> pixels.

## Workflow with multiple species

`id` and `species` can be defined using text files:

```{r}
load_common_data()
Spp <- read.table(system.file("extdata/species.txt", package="cure4insect"))
str(Spp)
ID <- read.table(system.file("extdata/pixels.txt", package="cure4insect"))
str(ID)
subset_common_data(id=ID, species=Spp)
xx <- report_all()
str(xx)
do.call(rbind, lapply(xx, flatten))
```

## Species subset specifications

Here is how to inspect all possible species IDs

```{r}
str(get_all_species())
str(get_species_table())
```

Select one or more taxonomic groups
(mammals, birds, mites, mosses, lichens, vpalnst),
and fiter for habitat and status:

```{r}
## birds and mammals
str(get_all_species(taxon=c("birds", "mammals")))
## all upland species
str(get_all_species(taxon="all", habitat="upland"))
## nonnative vascular plants
str(get_all_species(taxon="vplants", status="nonnative"))
```

## Wrapper functions

* `species="all"` runs all species
* `species="mites"` runs all mite species
* `sender="you@example.org"` will send an email with the results attached
* increase `cores` to allow parallel processing

```{r}
z <- custom_report(id=ID,
    species=c("AlderFlycatcher", "Achillea.millefolium"),
    address=NULL, cores=1)
z
```

Working with a local copy of the results is much faster
set path via function arguments or the options:

```{r}
## making of the file raw_all.rda
library(cure4insect)
opar <- set_options(path = "w:/reports")
getOption("cure4insect")
load_common_data()
subset_common_data(id=get_all_id(),
    species=get_all_species())
## see how these compare
system.time(res <- report_all(cores=1))
#system.time(res <- report_all(cores=2))
#system.time(res <- report_all(cores=4))
## this is for testing only
#system.time(res <- .report_all_by1())
(set_options(opar)) # reset options
```

A few more words about options:

```{r}
## options
getOption("cure4insect")
## change configs in this file to make it permanent for a given installation
as.list(drop(read.dcf(file=system.file("config/defaults.conf",
    package="cure4insect"))))
```

## Sector effects and intactness plots

```{r}
## *res*ults from calculate_results, all province, all species
load(system.file("extdata/raw_all.rda", package="cure4insect"))

plot_sector(res[["CanadaWarbler"]], "unit")
plot_sector(res[["CanadaWarbler"]], "regional")
plot_sector(res[["CanadaWarbler"]], "underhf")

z <- do.call(rbind, lapply(res, flatten))
class(z) <- c("c4idf", class(z))
plot_sector(z, "unit") # all species
plot_sector(z[1:100,], "regional") # use a subset
plot_sector(z, "underhf", method="hist") # binned version

plot_intactness(z, "SI")
plot_intactness(z, "SI2", method="hist")
```

## Determining spatial IDs based on spatial polygons

`id` can also be a SpatialPolygons object based on GeoJSON for example:

```{r}
library(rgdal)
dsn <- system.file("extdata/polygon.geojson", package="cure4insect")
cat(readLines(dsn), sep="\n")
ply <- readOGR(dsn=dsn)
subset_common_data(id=ply, species=Spp)
plot(make_subset_map())
xx2 <- report_all()
```

Spatial IDs of the 1 km<sup>2</sup> spatial pixel units are to be
used for the custom summaries.
The `Row_Col` field defines the IDs and links the raster cells to the [geodatabase](http://ftp.public.abmi.ca/species.abmi.ca/gis/Grid1km_working.gdb.zip)
or [CSV](http://ftp.public.abmi.ca/species.abmi.ca/gis/Grid1km_working.csv.zip}) (with latitude/longitude in [NAD_1983_10TM_AEP_Forest](http://spatialreference.org/ref/epsg/3402/) projection).

For the [web application](http://sc-dev.abmi.ca/ocpu/apps/ABbiodiversity/cure4insect/www/),
use your favourite GIS software, or in R use this to get the spatial IDs
written into a text file:

```{r}
library(rgdal)
load_common_data()
dsn <- system.file("extdata/OSA_bound.geojson", package="cure4insect")
ply <- readOGR(dsn=dsn)
ID <- overlay_polygon(ply)
## write IDs into a text file
write.table(data.frame(SpatialID=ID), row.names=FALSE, file="SpatialID.txt")

## spatial pixels: selection in red
xy <- get_id_locations()
plot(xy, col="grey", pch=".")
plot(xy[ID,], col="red", pch=".", add=TRUE)

## compare with the polygons
AB <- readOGR(dsn=system.file("extdata/AB_bound.geojson",
    package="cure4insect"))
plot(AB, col="grey")
plot(ply, col="red", add=TRUE)
```

Use the `make_subset_map()` function to get a raster map of
the spatial selection.

## Raster objects and maps

The result is a raster stack object with the following layers:

* `NC`, `NR`: current and reference abundance,
* `SI`, `SI2`: one- and two-sided intactness,
* `SE`, `CV`: bootstrap based standard error and coefficient of variation
estimates for current abundance.

```{r}
load_common_data()
y <- load_species_data("Ovenbird")
r <- rasterize_results(y)
plot(r, "NC") # current abundance map
col <- colorRampPalette(c("darkgreen","yellow","red"))(250)
plot(r, "SE", col=col) # standadr errors for current abundance
```

It is possible to make multi-species maps as well:
average intactness and expected number of species.

```{r}
subset_common_data(species=get_all_species(taxon="birds"))
r1 <- make_multispecies_map("richness")
r2 <- make_multispecies_map("intactness")
```

## Spatially explicit (polygon level) predictions

The 1 km<sup>2</sup> level predictions provide mean abundance per pixel.
Sometimes we need finer detail, e.g. when making predictions as part of
spatially explicit simulations.

First we load the spatial/climate related component of the predictions
(which is a raster object):

```{r}
load_common_data()
species <- "Achillea.millefolium"
object <- load_spclim_data(species)
```

The spatial component is then combined with the land cover component
describing vegetation/disturbance/soil classes as a factor.

```{r}
## original levels
levels(veg <- as.factor(get_levels()$veg))
levels(soil <- as.factor(get_levels()$soil))
```
Sometimes it is best to create a crosswalk table and
reclassify using e.g. the `mefa4::reclass` function:

```{r}
(rc <- data.frame(In=c("pine5", "decid15", "urban", "industrial"),
    Out=c("Pine0", "Deciduous10", "UrbInd", "UrbInd")))
mefa4::reclass(c("pine5", "pine5", "decid15", "urban", "industrial"), rc)
```

We need to have spatial locations for each land cover value
(same value can be repeated, but but avoid duplicate rownames).
We use the **sp** package to make a SpatialPoints object:

```{r}
XY <- get_id_locations()
coords <- coordinates(XY)[10^5,,drop=FALSE]
rownames(coords) <- NULL
xy <- data.frame(coords[rep(1, length(veg)),])
coordinates(xy) <- ~ POINT_X + POINT_Y
proj4string(xy) <- proj4string(XY)
```

Now we are ready to make the predictions:

```{r}
pred <- predict(object, xy=xy, veg=veg)
summary(pred)
```

The `predict` function returns a data frame with columns `veg`, `soil`, and
`comb` (combines `veg` and `soil` based on aspen probability of occurrence
using `combine_veg_soil` as a weighted average based on
probability of aspen occurrence).

For some species, either the `veg` or `soil` based estimates are unavailable:
`predict` returns `NA` for these and the combined results will be `NA` as well.

The next line is a more succinct version that loads the species data as well,
but we can't reuse the species data after:

```{r}
pred <- custom_predict(species, xy=xy, veg=veg)
```

Another was of making predictions is to define a spatial grid, and quantify
land cover as proportion of the land cover types in each grid cell.
This is how we can use multivariate input data in a spatial grid
(totally unrealistic data set just for illustration,
but user has to make sure the numbers are meaningful):

```{r}
xy <- xy[1:10,]
mveg <- matrix(0, 10, 8)
colnames(mveg) <- veg[c(1:8 * 10)]
mveg[] <- rpois(80, 10) * rbinom(80, 1, 0.2)
mveg[rowSums(mveg)==0,1] <- 1 # avoid 0 row sum
mveg

msoil <- matrix(0, 10, 6)
colnames(msoil) <- get_levels()$soil[1:6]
msoil[] <- rpois(60, 10) * rbinom(60, 1, 0.4)
msoil[rowSums(msoil)==0,1] <- 1 # avoid 0 row sum
msoil
```

Because we used areas (not proportions) we get the output as
two matrices containing abundances (density times area) corresdonding to the
vegetation and soil matrices:

```{r}
(prmat1 <- predict_mat(object, xy, mveg, msoil))
```

Row sums give the total abundance at each location,
column sums give the total abundance in a land cover type over all locations:

```{r}
rowSums(prmat1$veg)
colSums(prmat1$veg)
```

Using proportions in the input matrices gives mean abundance per spatial unit
as output:

```{r}
(prmat2 <- predict_mat(object, xy, mveg/rowSums(mveg), msoil/rowSums(msoil)))
```

Combining vegetation and soil based predictions returns a vector,
i.e. the aspen probability weighted average of the vegetation and soil
based total abundances:

```{r}
combine_veg_soil(xy, rowSums(prmat2$veg), rowSums(prmat2$soil))
```

## Visualize land cover associations

See the following [R markdown](http://rmarkdown.rstudio.com/)
file for a worked example of visualizations available in the package:

```{r}
file.show(system.file("doc/example-species-report.Rmd", package="cure4insect"))
```

Habitat associations as shown on the [species.abmi.ca](http://species.abmi.ca/) website:

```{r}
load_common_data()
plot_abundance("Achillea.millefolium", "veg_coef")
plot_abundance("Achillea.millefolium", "soil_coef", paspen=1)
plot_abundance("Achillea.millefolium", "veg_lin")
plot_abundance("Achillea.millefolium", "soil_lin")
```

## Web API

The web app sits [here](http://sc-dev.abmi.ca/ocpu/apps/ABbiodiversity/cure4insect/www/).
To get more control over the results, use the [API](https://www.opencpu.org/api.html#api-formats).

Make a request using the `custom_report` function:

```shell
curl http://sc-dev.abmi.ca/ocpu/apps/ABbiodiversity/cure4insect/R/custom_report/csv \
-H "Content-Type: application/json" -d \
'{"id":["182_362", "182_363"], "species":["AlderFlycatcher", "Achillea.millefolium"]}'
```

Access spatially explicit and land cover specific prediction for a species
using the `custom_predict` function:

```shell
curl http://sc-dev.abmi.ca/ocpu/apps/ABbiodiversity/cure4insect/R/custom_predict/json \
-H "Content-Type: application/json" -d \
'{"species":"AlderFlycatcher", "xy":[[-114.4493,58.4651]], "veg":"Mixedwood80"}'
```

## Explore single and multi-species results

To get similar output to
[this](https://abbiodiversity.github.io/cure4insect/site/),
run script from this file:

```{r}
file.show(system.file("doc/custom-report.R", package="cure4insect"))
```

